Gut-resident microorganisms and their genes are associated with cognition and neuroanatomy in children

### 2023-10-11 manuscript/main.tex
remove highlights

Given that the development of different domains of cognitive function
(e.g., gross motor development vs visual perception) may develop at different times
and be affected by different molecular signals,
we sought to determine whether the associations of microbial taxa
with cognitive function are general across all cognitive subdomains.
To accomplish this, we assessed individual subscales of the  
Mullen Scales of Early Learning (MSEL, birth to 36 months - \textit{see Methods}). 
Taxonomic profiles coupled with demographics exhibited predictive potential towards three
of five probed subscales (Figure 4A), 
namely Expressive Language ( test-set RMSE \hl{12.05}, test-set correlation \hl{0.368}), 
Gross Motor (test-set RMSE \hl{11.07}, test-set correlation \hl{0.137}) and 
Visual Reception (test-set RMSE \hl{15.2}, test-set correlation \hl{0.333}).
Between the subscales, not only the predictability (measured by the test-set figures of merit) is different,
but also the relative importances of several taxa (Figure 4B).
Although certain species such as \textit{B. longum}, \textit{F. prausznitzii} 
have similar relative importance across all subscales, others are substantially different.
Notably, \textit{Bifidobacterium pseudocatenulatum}, \textit{B. wexlerae} and \textit{E. eligens} are
substantially more important for the prediction of EL, while
\textit{R. faecis}, \textit{Streptococcus salivarius} and \textit{F. saccharivorans} are
distinctly predictive of GM, and
\textit{Clostridium innocuum}, \textit{B. vulgatus} 
stand out on models targeting visual reception.

@@

\emph{A. celatus}, an equol-producing species, is an examples of the second contribution
pattern, where a taxon only contributes significantly to a very small number of segments. 
This species was highly important in predicting the relative volume of
the right anterior cingulate (\hl{2.8\%} relative importance).
Another example of this pattern is \emph{Coprococcus eutactus}
which is heavily associated with the left precentral area (\hl{3.5\%} importance). 
Finally, \emph{C. comes} has an importance
distribution that splits almost evenly among the two previously reported
clusters. Its loading on the prediction of the left posterior cingulate
is the highest for a microbe in RF models (\hl{3.7\%} importance);
only age had higher relative importance in any model. At the same time,
\textit{C. comes} is one of the most important predictors for neighboring areas of the
posterior cingulate such as the pars opercularis (importances,
\hl{Left = 1.9\%, Right = 2.5\%}), along with upper regions like the left
precentral (2.0\% importance) and paracentral lobes (importances, \hl{Left = 2.0\%, Right = 1.5\%}).

>>>>

Given that the development of different domains of cognitive function
(e.g., gross motor development vs visual perception) may develop at different times
and be affected by different molecular signals,
we sought to determine whether the associations of microbial taxa
with cognitive function are general across all cognitive subdomains.
To accomplish this, we assessed individual subscales of the  
Mullen Scales of Early Learning (MSEL, birth to 36 months - \textit{see Methods}). 
Taxonomic profiles coupled with demographics exhibited predictive potential towards three
of five probed subscales (Figure 4A), 
namely Expressive Language ( test-set RMSE 12.05, test-set correlation 0.368), 
Gross Motor (test-set RMSE 11.07, test-set correlation 0.137) and 
Visual Reception (test-set RMSE 15.2, test-set correlation 0.333).
Between the subscales, not only the predictability (measured by the test-set figures of merit) is different,
but also the relative importances of several taxa (Figure 4B).
Although certain species such as \textit{B. longum}, \textit{F. prausznitzii} 
have similar relative importance across all subscales, others are substantially different.
Notably, \textit{Bifidobacterium pseudocatenulatum}, \textit{B. wexlerae} and \textit{E. eligens} are
substantially more important for the prediction of EL, while
\textit{R. faecis}, \textit{Streptococcus salivarius} and \textit{F. saccharivorans} are
distinctly predictive of GM, and
\textit{Clostridium innocuum}, \textit{B. vulgatus} 
stand out on models targeting visual reception.

@@

\emph{A. celatus}, an equol-producing species, is an examples of the second contribution
pattern, where a taxon only contributes significantly to a very small number of segments. 
This species was highly important in predicting the relative volume of
the right anterior cingulate (2.8\% relative importance).
Another example of this pattern is \emph{Coprococcus eutactus}
which is heavily associated with the left precentral area (3.5\% importance). 
Finally, \emph{C. comes} has an importance
distribution that splits almost evenly among the two previously reported
clusters. Its loading on the prediction of the left posterior cingulate
is the highest for a microbe in RF models (3.7\% importance);
only age had higher relative importance in any model. At the same time,
\textit{C. comes} is one of the most important predictors for neighboring areas of the
posterior cingulate such as the pars opercularis (importances,
Left = 1.9\%, Right = 2.5\%), along with upper regions like the left
precentral (2.0\% importance) and paracentral lobes (importances, Left = 2.0\%, Right = 1.5\%).
